Global multivariate model learning from hierarchically correlated data

### 2021-04-26 revision/response_to_referees.tex

%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{bm}% bold math

\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{response}{rgb}{0.1, 0.1, 0.1}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
\newcommand{\refa}[1]{\textbf{R1:} #1\vskip 5mm}
\newcommand{\refb}[1]{\textbf{R2:} #1\vskip 5mm}
\newcommand{\refc}[1]{\textbf{R3:} #1\vskip 5mm}

\newcommand{\editor}[1]{\textbf{Editor:} #1\vskip 5mm}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
\newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}

\begin{document}
\section*{Response to reviewers}

\subsection*{Editor}

\editor{}

\response{}

\subsection*{Reviewer: 1}

\refa{}

\response{}

\subsection*{Reviewer: 2}

\refb{
}

\response{}

\subsection*{Reviewer: 3 - Main text}

\refc{I think the authors should add some comparison to other approaches of inference for OU processes (for example in the introduction), to allow the reader to understand how the work sits in the context of dynamical inference methods}
\responsedraft{Link with comments of Reviewer 1: we can probably group the two answers and modify the intro in a way that satisfies both.}

\refc{I think that ‘statement of the problem’ is an important part that probably deserves to be a section in itself instead of a subsection of ‘Methods’.}
\response{``Statement of the problem'' is now a section in itself.}

\refc{The extreme cases discussed at lines 193-195 seem the same discussed at lines 144-147. If this is the case, the authors could re-think whether it is worth to repeat this point or to establish explicitly the connection to what has been discussed before.}
\response{We agree with reviewer 3 that this was redundant, and modified the text accordingly. 
The last paragraph of section II (former lines 144-147) now introduces timescales of the OU process without referring to the inference procedure. 
Former lines 193-195 then detail what these timescales imply for the inference procedure (the two extreme cases), referring to section II.}

\refc{Is it necessary to introduce a new symbol, $z$, for the expression (13)? 
Note that $z$ is used also in the Appendix for something with a different meaning (e.g. line 637)}
\response{The symbol $z$ is only used in the couple of equations that follow its introduction, and we agree that it is not strictly necessary. 
However, we felt that it made the derivation clearer by separating the calculation of eigenvalues and eigenvectors of $\mathbb{G}$ from the specific functional form of eigenvalues of $\mathbf{\Lambda}$.}

\refc{I think it is not immediately clear what the ‘Results’ section will address. I think one could immediately state (like at line 316) that the goal is to assess the accuracy of Cmax compared to the real C and that such estimate will be compared to the alternativa estimate provided by the empirical C. The definition of empirical C could be also explicitly written down. Later on, at line 374, the authors mention the ‘empirical coupling matrix’ but it is not clear how it is derived, is this simply then the inverse of the empirical C? Clarifying this would help better understand why the fact that C is close to singular (line 349) has an impact especially on the error on J (Fig. 3 right).}
\response{We agree with the reviewer that the Results section lacked explanation. 
We added an introductory paragraph that states the goal (evaluating the accuracy of the inference procedure) and describes how we proceed.
The empirical covariance matrix as well as the different coupling matrices are also clearly defined in this new paragraph. }

\refc{I thought also that it is not clear how the predictions for an ‘i.i.d sample’ (lines 355, 376) are obtained: do they serve as a random control giving the random expectation in for the errors in e.g. Fig. 3? I think this should be stated more clearly, also in the caption, maybe along with the numerical value itself of the i.i.d. error (that is difficult to read from the plots’s scale).}
\responsedraft{{\bf{Edwin}: any way that you can get and add to the text the exact numerical values for the error of the i.i.d. of sizes 16 / 32 / 64? Those are the examples we give in the text.}}
\response{The predictions of the \emph{i.i.d.} sample can indeed be used as a control to compute the minimal error that we can expect for a sample size $N$. 
Here, we use them to define an effective sample size: to each value of the  reconstruction error $\vert\vert \bm{C}_{max} - \bm{C}\vert\vert_2$ corresponds an \emph{i.i.d.} sample of size $N$ for which the same $l2$-error is obtained when using the empirical estimation of the covariance matrix.
This allows us to formulate the gains obtained by our method in terms of an equivalent increase in \emph{i.i.d.} sample size.

We have reformulated the corresponding paragraph to clarify this idea, and added more explanations to the caption of figure S5.}

\refc{Line 336: It is written that the sampling is repeated 100 times, it is not clear if also the inference is repeated 100 times or if this is done to increase $N$, the sample size. In relation to this, it would be important to write somewhere what $N$ is considered in the computational tests and hence what is the computational complexity given this $N$ (and $L=4,10$).}
\responsedraft{What does he mean by computational complexity given values of $N$ and $L$? 
I've only ever seen computational complexity expressed as an asymptotic quantity.}
\response{The inference is indeed repeated a hundred times. 
We do this to average our results accross different realizations of the OU process. 
We now clearly state $N$ corresponds to the number of leaves of the tree, in our case $2^9=512$.}

\refc{Line 370: Please add some references for the problem of contacts in proteins.}
\refc{Line 421: Please add some references for the evidence that protein landscapes are well described by pairwise models.}
\responsedraft{TODO}

\refc{Regarding the results, do the authors believe that inferring $\gamma$ could be interesting in terms of biological applications? Could it inform about typical evolutionary timescales?}
\response{The inference of $\gamma$ indeed provides information about typicaly times of the underlying evolutionary process, provided that the eigenvalues of the covariance matrix $\bm{C}$ are also known (Eq.~(7) and comments below it). 
If the evolution of, say, protein sequences is reasonably well described by the OU process, we could then be able to make statements about the expected change in an amino acid sequence as a function of time or number of generations.

A current limitation of that approach is that $\gamma$ will be inferred using distances separating leaves on a tree. 
In typical tree-inference software from biological sequences, the branch length is related to the expected number of mutations on that branch, but cannot be directly related to physical time or number of generations. 
}
\responsedraft{Not sure if this is a good answer. Branch lengths on trees can be related to real time using a molecular clock hypothesis, but it's unclear to me if that's consistent with the OU process itself?}

\refc{Fig. 3: I do not understand what the legend symbol for C_iid refers to (left panel). Is this simply the dotted horizontal line at the bottom of the plot? If yes I do not see why the legend has an additional central dot. In the caption, I do not understand ‘The inset in both panels’, isn’t the inset only in the right panel? Same comment applied to caption of Fig. S9.}
\responsedraft{\bm{Edwin}: Can you change the legend to reflect what's on the figure? Legend for Ciid should be a dashed line only. }
\response{The reviewer is right: the legend for figures 3 and 9 is incorrect -- it should be a dashed line only for $\bm{C}_{iid}$ and $\bm{J}_{iid}$ -- and the inset figure is only present in the right panel. 
We have corrected these errors. }

\refc{
	\pierre{Below: issues that I have addressed by changing the MS, but that perhaps do not require one response each.}
	\begin{itemize}
		\item Line 191: I would state explicitly why the inference becomes impossible
		\item Line 240: should ``we construct the direct product of vectors $s_{a}$ and $u_{ka}$, defining vectors $S_{ka}$ of dimensions $L \times N$’ be ‘we construct the direct product of vectors $u_{ka}$ and $s_{a}$, defining vectors $S_{ka}$ of dimensions $N \times L$''?
		I would also state explicitly that the superindex refers to the components of $u$.
		\item Line 278: double brackets around 17.
		\item Is there a minus in front of r.h.s of (19)?
		\item Line 332: I would add a comment on the choice of the range for gamma (why not even smaller or bigger) - this is probably due to the fact that all the variation concentrates in the range considered but I would state it.
		\item Line 346: As far as I see the l2-error is not defined anywhere - I think it would be important to have it defined somewhere (could be also in a caption or in the appendix).
		\item Fig 4: I think the caption should state clearly what the ‘Perfect’ case refers to. Same comment applies to caption of Fig. S10.
		\item I think the author of Ref. [15] was repeated twice by mistake.
	\end{itemize}
}

\end{document}

### 2021-04-26 main.tex
answer comments of R3 on main text

One of the biggest application areas of inverse statistical mechanics is the modeling of biological processes.
These applications are fuelled by the large amount of available data resulting from the impressive progress in experimental techniques in biology. 
This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018}.
A common idea in this context is that it is possible to use characteristics of genes or organisms related by a common ancestry -- called \emph{homologous} -- to construct models of the selection acting on them. 
A successful example in this regard is the representation of protein sequences by probabilistic models in the so-called DCA method \cite{levy_potts_2017,cocco_inverse_2018}. 
The prototypical datasets in this context are multiple-sequence alignments (MSA), with lines being a so-called homologous, {\em i.e.} evolutionarily related sequences, and columns specific positions deriving from some common ancestral position \cite{durbin1998biological}.  
The MSA contains at least two kinds complementary information:
\begin{itemize}
    \item {\it Phylogenetic information:} the distances between sequences carry information about the evolutionary time since their common ancestor. Using phylogeny inference methods \cite{felsenstein_phylogenies_1988,felsenstein2004inferring} we may reconstruct the evolutionary history of our dataset, represented by a phylogenetic tree.  
    \item {\it Co-evolutionary information:} positions in a sequence typically do not evolve independently, but rather in a correlated way. This co-evolution carries important information about the selection forces acting on evolving entities. This fact has been extensively studied in the case of protein sequences, and used to predict structure, mutational landscapes or networks of interacting proteins \cite{levy_potts_2017,cocco_inverse_2018}.  
\end{itemize}
These two types of information are contained in two complementary features of the data: phylogenetic inference is based on the comparative analysis of different sequences, while co-evolutionary information is contained in the correlation of different columns of the MSA. 
Modeling approaches using one type of information typically neglect the other one: inference of phylogenies generally assumes that all positions in a sequence evolve independently, while co-evolutionary models of proteins assume that sequences in the MSA are independently distributed. 
This choice is motivated by the fact that taking the two types of correlations into account, \emph{i.e.} through time with phylogeny and accross trait values for co-evolution, results in very hard inference procedures, cf.~\cite{obermayer2014inverse,rodriguez2019toward}. 
However, this can lead to biases in the model parameters: it has for instance been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function \cite{qin_power_2018,horta2020phylogenetic}.

@@

In this work, we are interested in constructing an inference method for parameters of an OU process from data correlated through a tree.
Our approach is purely methodological, and the data can represent any set of continuous phenotypic traits, \emph{e.g.} from different organisms, with the tree indicating the phylogenetic relations between data points. 
Inferred parameters then represent the selection forces without biasing effects from the phylogeny. 
The manuscript is divided as follows: we first review in section \ref{sec:ornstein_uhlenbeck_dynamics} the main characteristics of the multivariate OU process. 
We then describe the setting of the inference problem that we want to solve in section \ref{sub:statement_of_the_problem}, propose a solution in sections \ref{sub:Calculation_of_the_likelihood} and \ref{sub:maximization_of_the_likelihood}.
Finally, we present results obtained on simulated data in section \ref{sec:Results}, with the context of pairwise models of protein sequences in mind.

@@

Eq.~\eqref{eq:pairwisecov} allows us to readily distinguish two regimes. 
Let us call $\rho_a$ the eigenvalues of $\bm{J}$. 
The eigenvalues of $\Lam^{\Delta t}\bm{C}$ are then equal to $\rho_a^{-1}e^{-\gamma\rho_a\Delta t}$. 
Since all $\rho_a$ are positive, the eigenvalues of $\Lam^{\Delta t}\bm{C}$ vanish exponentially over time. The slowest timescale of exponential decay is set by $\tau_c^{-1} = \gamma\rho_{min}$, with $\rho_{min}$ being the smallest eigenvalue of $\bm{J}$. 
Thus, for $\Delta t / \tau_c \gg 1$, $\vx_1$ and $\vx_2$ are uncorrelated. 
If this is verified for all pairs of observations $\vx_i$ and $\vx_j$, the regime is that of \emph{uncorrelated} data -- the inference of $\bm{J}$ can simply be performed by inverting the empirical covariance matrix extracted from the data. 
Inversely, for $\Delta t / \tau_c \ll 1$, $\vx_1$ and $\vx_2$ are highly correlated, defining a \emph{strongly correlated} regime. 
It should be noted that for $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and Eq.~\eqref{eq:pairwisecov} becomes irrelevant. Actually, $\vx_1$ and $\vx_2$ coincide at that point, \ie~we have $P(\vx_1,\vx_2|\,\Delta t=0) = P_{eq}(\vx_1) \times \delta(\vx_1 - \vx_2)$ using the $L$-dimensional Dirac distribution.

\subsection{Statement of the problem}
\label{sub:statement_of_the_problem}

@@

This problem shows two notable extreme cases: 
The first one is the case where the typical branch length of the tree is short compared to the timescales of the OU process.
As a consequence, leaf configurations are close to identical to the root, \ie~$\vx_i\simeq\vx_r$, and the inference of $\bm C$ becomes impossible. 
The second one is the opposite case where the typical branch length of the tree is long compared to the longest timescale of the OU process $\tau_c$.
In this case, the configuration of a child node is close to independent from that of its ancestor, and leaf configurations can be considered as independent samples from the equilibrium distribution $P_{eq}$.
$\bm C$ can then be readily estimated by computing the empirical covariance matrix. 
We are interested here in the intermediate regime where substantial tree-mediated correlations between data
make it impossible to simply estimate $\bm C$ with the empirical covariance, but the depth of the tree introduces enough variability in the data for one to hope of reconstructing the energy potential $\bm{J}$.

@@

To obtain the eigenmodes of the joint covariance matrix $\mathbb{G}$ as a function of the $\vsa$ and $\vuka$, we construct the direct product of vectors $\vsa$ and $\vuka$, defining vectors $\vec{S}_{ka}$ of dimension $L\times N$:
\begin{equation}
	\begin{split}
		\vec{S}_{ka} &= \vuka \otimes \vsa \\
		&= [u_{ka}^1\cdot\vsa, \ldots, u_{ka}^N\cdot\vsa].
		% &= [u_{ka}(1)s_a(1), \ldots, u_{ka}(1)s_a(L); \ldots; u_{ka}(N)s_a(1), \ldots, u_{ka}(N)s_a(L)]
	\end{split}
	\label{eq:Sdef}
\end{equation}
The $i$th block vector of $\vec{S}_{ka}$ will thus be written as $\vec{S}_{ka}^i = u_{ka}^i\cdot\vsa$. 
We can now show that $\vec{S}_{ka}$ are eigenvectors of matrix $\mathbb{G}$ by considering the $i$th block vector of the product $\mathbb{G}\cdot\vec{S}_{ka}$:
\begin{equation}
	\begin{split}
		\left(\mathbb{G}\cdot\vec{S}_{ka}\right)^i &= \sum_{j=1}^N \mathbb{G}_{ij} u_{ka}^j\cdot\vsa \\
		&= \sum_{j=1}^N z(\rho_a, \Delta t_{ij}) u_{ka}^j\cdot \vsa \\
		&= (\mathbf{G}^a\cdot\vuka)^i \cdot \vsa\\
		&= \lambda_{ka}(u_{ka}^i\cdot\vsa) \\
		&= \lambda_{ka}\vec{S}_{ka}^i \ .
	\end{split}
\end{equation}
We have first used the fact that $\vsa$ is an eigenvector of $\mathbb{G}_{ij}$, then the definition of ${\bm G}^a$, and finally the fact that $\vuka$ is an eigenvector of ${\bm G}^a$. 
This demonstrates that the eigenmodes of $\mathbb{G}$ are $\left\{ \lambda_{ka}, \vec{S}_{ka} \right\}$ with $1\leq k\leq N$ and $1\leq a \leq L$. 
Since $\mathbb{G}$ is the covariance matrix of a Gaussian distribution, we conclude the $\lambda_{ka}$ to be strictly positive.
Interestingly, the definition of $\vec{S}_{ka}$ as a direct product between eigenvectors $\vsa$ of the energy potential and eigenvectors $\vuka$ reflecting the correlation structure mediated by the tree illustrates how these two types of information are entangled in the covariance matrix of the data.

@@

As stated at the beginning of this section, our main task is to find the equilibrium covariance matrix $\bm C$ that maximizes the likelihood of the data. 
We also need to find the optimal time scale $\gamma$. 
In Eq.~(\eqref{eq:likelihood_3}), the likelihood is expressed as a function of $\gamma$ and $\{\rho_a, \vsa\}$, \ie~the eigenvalues and eigenvectors of $\iC$, either directly or through the quantities $\{\lambda_{ka}, \vuka \}$. 
We know attempt to maximize the likelihood with respect to the eigenmodes $\{\rho_a, \vsa\}$ and to the time scale $\gamma$.

@@

The gradient of the likelihood is obtained by differentiating Eq.~\eqref{eq:likelihood_3} with respect to the parameters of interest. 
This gives us  three equations: 
% dL / d rho
\begin{equation}
	\begin{split}
		\frac{\partial \mathcal{L}}{\partial\rho_a} &= -\frac 1 2 \sum_{k=1}^N \left\{ \frac{\partial\lambda_k}{\partial\rho_a}\lambda_k^{-1} - \frac{\partial\lambda_k}{\partial\rho_a} \lambda_k^{-2}\left(\sum_{i=1}^N u_{k}^i\vx_i\cdot\vsa\right)^2 \right. \\
		&+ \left. 2\lambda_k^{-1} \left(\sum_{i=1}^N u_k^i\vx_i\cdot\vsa\right) \left(\sum_{i=1}^N \frac{\partial u_k^i}{\partial\rho_a} \vx_i\cdot\vsa\right) \right\}, 
	\end{split}
	\label{eq:gradlikelihood_1}
\end{equation}
% dL / d theta
\begin{equation}
	\frac{\partial \mathcal{L}}{\partial\theta_{\alpha\beta}} = \sum_{k=1}^N \lambda_k^{-1} \left(\sum_{i=1}^N u_k^i\vx_i\cdot\vsa\right) \left(\sum_{i=1}^N u_k^i \vx_i\cdot\frac{\partial \vsa}{\partial\theta_{\alpha\beta}}\right),
	\label{eq:gradlikelihood_2}
\end{equation}
and
% dL / d gamma
\begin{equation}
	\begin{split}
		\frac{\partial \mathcal{L}}{\partial\gamma} &= -\frac 1 2 \sum_{k=1}^N \left\{ \frac{\partial\lambda_k}{\partial\gamma}\lambda_k^{-1} - \frac{\partial\lambda_k}{\partial\gamma} \lambda_k^{-2}\left(\sum_{i=1}^N u_{k}^i\vx_i\cdot\vsa\right)^2 \right. \\
		&+ \left. 2\lambda_k^{-1} \left(\sum_{i=1}^N u_k^i\vx_i\cdot\vsa\right) \left(\sum_{i=1}^N \frac{\partial u_k^i}{\partial\gamma} \vx_i\cdot\vsa\right) \right\}, 
	\end{split}
	\label{eq:gradlikelihood_3}
\end{equation}
The derivatives of $\vec{u}_k(\rho,\gamma)$ and $\lambda_k(\rho,\gamma)$ with respect to $\rho$ can then be computed using the following equations~\cite{matrix_cook_book}: 
\begin{equation}
	\frac{\partial\lambda_i(\rho,\gamma)}{\partial\rho} = \vec{u}_k(\rho,\gamma)^T\frac{\partial\bm{G}(\rho,\gamma)}{\partial\rho}\vec{u}_k(\rho,\gamma)
	\label{eq:grad_eigval}
\end{equation}
and
\begin{equation}
	\frac{\partial \vec{u}_k(\rho,\gamma)}{\partial\rho} = \sum_{l\neq k} \left( \vec{u}_k(\rho,\gamma)^T\frac{\partial\bm{G}(\rho,\gamma)}{\partial\rho}\vec{u}_l(\rho,\gamma) \right) \left(\lambda_k(\rho,\gamma) - \lambda_l(\rho,\gamma)\right)^{-1} \vec{u}_l(\rho,\gamma).
	\label{eq:grad_eigvec}
\end{equation}
Equivalent equations can be written for their derivatives with respect to $\gamma$.

@@

In order to evaluate our inference procedure, we generate artificial data corresponding to the process described in section \ref{sub:statement_of_the_problem}. 
We first build a  balanced binary tree $\mathcal{T}$ with $2^9=512$ leaves. 
The length of each branch of $\mathcal{T}$ is chosen from a uniform distribution in the interval $[0,1]$. 
We also sample positive semi-definite coupling matrix $\bm{J}$ of size $L\times L$ with $L=4$ or $L=10$, with entries normally distributed with mean $\mu_J=0.8$ and $\sigma_J=0.2$.

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/PPVs_L4.pdf}
	\caption{Quality of prediction of interactions for  different values of $\gamma$ and system size $L=4$. Interactions are defined as non-zero elements of the coupling matrix. In the $L=4$ case, there are $6$ possible interactions. Predictions are made by taking the largest elements (in absolute terms) of the inferred coupling matrix. The PPV is the fraction of correctly predicted contacts for a given number of predictions.}
	\label{fig:PPV_L4}
\end{figure}

@@

\clearpage
\bibliography{bib_phylo}
\bibliographystyle{unsrt}

>>>>

One of the biggest application areas of inverse statistical mechanics is the modeling of biological processes.
These applications are fuelled by the large amount of available data resulting from the impressive progress in experimental techniques in biology. 
This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018}.
A common idea in this context is that it is possible to use characteristics of genes or organisms related by a common ancestry -- called \emph{homologous} -- to construct models of the selection acting on them. 
A successful example in this regard is the representation of protein sequences by probabilistic models in the so-called DCA method \cite{levy_potts_2017,cocco_inverse_2018}. 
The prototypical datasets in this context are multiple-sequence alignments (MSA), with lines being a so-called homologous, {\em i.e.} evolutionarily related sequences, and columns specific positions deriving from some common ancestral position \cite{durbin1998biological}.  
The MSA contains at least two kinds complementary information:
\begin{itemize}
    \item {\it Phylogenetic information:} the distances between sequences carry information about the evolutionary time since their common ancestor. Using phylogeny inference methods \cite{felsenstein_phylogenies_1988,felsenstein_joseph_inferring_2003} we may reconstruct the evolutionary history of our dataset, represented by a phylogenetic tree.  
    \item {\it Co-evolutionary information:} positions in a sequence typically do not evolve independently, but rather in a correlated way. This co-evolution carries important information about the selection forces acting on evolving entities. This fact has been extensively studied in the case of protein sequences, and used to predict structure, mutational landscapes or networks of interacting proteins \cite{levy_potts_2017,cocco_inverse_2018}.  
\end{itemize}
These two types of information are contained in two complementary features of the data: phylogenetic inference is based on the comparative analysis of different sequences, while co-evolutionary information is contained in the correlation of different columns of the MSA. 
Modeling approaches using one type of information typically neglect the other one: inference of phylogenies generally assumes that all positions in a sequence evolve independently, while co-evolutionary models of proteins assume that sequences in the MSA are independently distributed. 
This choice is motivated by the fact that taking the two types of correlations into account, \emph{i.e.} through time with phylogeny and accross trait values for co-evolution, results in very hard inference procedures, cf.~\cite{obermayer2014inverse,rodriguez2019toward}. 
However, this can lead to biases in the model parameters: it has for instance been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function \cite{qin_power_2018,horta2020phylogenetic}.

@@

In this work, we are interested in constructing an inference method for parameters of an OU process from data correlated through a tree.
Our approach is purely methodological, and the data can represent any set of continuous phenotypic traits, \emph{e.g.} from different organisms, with the tree indicating the phylogenetic relations between data points. 
Inferred parameters then represent the selection forces without biasing effects from the phylogeny. 
The manuscript is divided as follows: we first review in section \ref{sec:ornstein_uhlenbeck_dynamics} the main characteristics of the multivariate OU process. 
We then describe the setting of the inference problem that we want to solve in section \ref{sec:statement_of_the_problem} and propose a solution in section \ref{sec:methods}.
Finally, we present results obtained on simulated data in section \ref{sec:Results}, with the context of pairwise models of protein sequences in mind.

@@

Eq.~\eqref{eq:pairwisecov} allows us to identify typical timescales associated to the OU process. 
Let us call $\rho_a$ the eigenvalues of $\bm{J}$. 
The eigenvalues of $\Lam^{\Delta t}\bm{C}$ are then equal to $\rho_a^{-1}e^{-\gamma\rho_a\Delta t}$. 
Since all $\rho_a$ are positive, the eigenvalues of $\Lam^{\Delta t}\bm{C}$ vanish exponentially over time over timescales of the form $\rho_a\Delta t$. 
The slowest timescale of exponential decay is set by $\tau_c^{-1} = \gamma\rho_{min}$, with $\rho_{min}$ being the smallest eigenvalue of $\bm{J}$. 
Thus, for $\Delta t / \tau_c \gg 1$, $\vx_1$ and $\vx_2$ are uncorrelated, whereas they are highly correlated for $\Delta t / \tau_c \ll 1$, $\vx_1$ and $\vx_2$.
It should be noted that for $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and Eq.~\eqref{eq:pairwisecov} becomes irrelevant. Actually, $\vx_1$ and $\vx_2$ coincide at that point, \ie~we have $P(\vx_1,\vx_2|\,\Delta t=0) = P_{eq}(\vx_1) \times \delta(\vx_1 - \vx_2)$ using the $L$-dimensional Dirac distribution.

\section{Statement of the problem}
\label{sec:statement_of_the_problem}

@@

This problem shows two notable extreme cases: 
The first one is the case where the typical branch length of the tree is short compared to the timescales introduced in section \ref{sec:ornstein_uhlenbeck_dynamics}
As a consequence, leaf configurations are close to identical to the root, \ie~$\vx_i\simeq\vx_r$, and the inference of $\bm C$ becomes impossible due to lack of independent samples from $P_{eq}$. 
The second one is the opposite case where the typical branch length of the tree is long compared to the longest timescale of the OU process $\tau_c$.
In this case, the configuration of a child node is close to independent from that of its ancestor, and leaf configurations can be considered as independent samples from the equilibrium distribution $P_{eq}$.
$\bm C$ can then be readily estimated by computing the empirical covariance matrix. 
We are interested here in the intermediate regime where substantial tree-mediated correlations between data
make it impossible to simply estimate $\bm C$ with the empirical covariance, but the depth of the tree introduces enough variability in the data for one to hope of reconstructing the energy potential $\bm{J}$.

@@

To obtain the eigenmodes of the joint covariance matrix $\mathbb{G}$ as a function of the $\vsa$ and $\vuka$, we construct the direct product of vectors $\vuka$ and $\vsa$, defining vectors $\vec{S}_{ka}$ of dimension $L\cdot N$:
\begin{equation}
	\begin{split}
		\vec{S}_{ka} &= \vuka \otimes \vsa \\
		&= [u_{ka}^1\cdot\vsa, \ldots, u_{ka}^N\cdot\vsa],
	\end{split}
	\label{eq:Sdef}
\end{equation}
where the superindex $i\in\{1\ldots N\}$ refers to components of $\vuka$.
The $i$th block vector of $\vec{S}_{ka}$ will thus be written as $\vec{S}_{ka}^i = u_{ka}^i\cdot\vsa$. 
We can now show that $\vec{S}_{ka}$ are eigenvectors of matrix $\mathbb{G}$ by considering the $i$th block vector of the product $\mathbb{G}\cdot\vec{S}_{ka}$:
\begin{equation}
	\begin{split}
		\left(\mathbb{G}\cdot\vec{S}_{ka}\right)^i &= \sum_{j=1}^N \mathbb{G}_{ij} u_{ka}^j\cdot\vsa \\
		&= \sum_{j=1}^N z(\rho_a, \Delta t_{ij}) u_{ka}^j\cdot \vsa \\
		&= (\mathbf{G}^a\cdot\vuka)^i \cdot \vsa\\
		&= \lambda_{ka}(u_{ka}^i\cdot\vsa) \\
		&= \lambda_{ka}\vec{S}_{ka}^i \ .
	\end{split}
\end{equation}
We have first used the fact that $\vsa$ is an eigenvector of $\mathbb{G}_{ij}$, then the definition of ${\bm G}^a$, and finally the fact that $\vuka$ is an eigenvector of ${\bm G}^a$. 
This demonstrates that the eigenmodes of $\mathbb{G}$ are $\left\{ \lambda_{ka}, \vec{S}_{ka} \right\}$ with $1\leq k\leq N$ and $1\leq a \leq L$. 
Since $\mathbb{G}$ is the covariance matrix of a Gaussian distribution, we conclude the $\lambda_{ka}$ to be strictly positive.
Interestingly, the definition of $\vec{S}_{ka}$ as a direct product between eigenvectors $\vsa$ of the energy potential and eigenvectors $\vuka$ reflecting the correlation structure mediated by the tree illustrates how these two types of information are entangled in the covariance matrix of the data.

@@

As stated at the beginning of this section, our main task is to find the equilibrium covariance matrix $\bm C$ that maximizes the likelihood of the data. 
We also need to find the optimal time scale $\gamma$. 
In Eq.~\eqref{eq:likelihood_3}, the likelihood is expressed as a function of $\gamma$ and $\{\rho_a, \vsa\}$, \ie~the eigenvalues and eigenvectors of $\iC$, either directly or through the quantities $\{\lambda_{ka}, \vuka \}$. 
We know attempt to maximize the likelihood with respect to the eigenmodes $\{\rho_a, \vsa\}$ and to the time scale $\gamma$.

@@

The gradient of the likelihood is obtained by differentiating Eq.~\eqref{eq:likelihood_3} with respect to the parameters of interest. 
This gives us  three equations: 
% dL / d rho
\begin{equation}
	\begin{split}
		\frac{\partial \mathcal{L}}{\partial\rho_a} &= -\frac 1 2 \sum_{k=1}^N \left\{ \frac{\partial\lambda_k}{\partial\rho_a}\lambda_k^{-1} - \frac{\partial\lambda_k}{\partial\rho_a} \lambda_k^{-2}\left(\sum_{i=1}^N u_{k}^i\vx_i\cdot\vsa\right)^2 \right. \\
		&+ \left. 2\lambda_k^{-1} \left(\sum_{i=1}^N u_k^i\vx_i\cdot\vsa\right) \left(\sum_{i=1}^N \frac{\partial u_k^i}{\partial\rho_a} \vx_i\cdot\vsa\right) \right\}, 
	\end{split}
	\label{eq:gradlikelihood_1}
\end{equation}
% dL / d theta
\begin{equation}
	\frac{\partial \mathcal{L}}{\partial\theta_{\alpha\beta}} = -\sum_{k=1}^N \lambda_k^{-1} \left(\sum_{i=1}^N u_k^i\vx_i\cdot\vsa\right) \left(\sum_{i=1}^N u_k^i \vx_i\cdot\frac{\partial \vsa}{\partial\theta_{\alpha\beta}}\right),
	\label{eq:gradlikelihood_2}
\end{equation}
and
% dL / d gamma
\begin{equation}
	\begin{split}
		\frac{\partial \mathcal{L}}{\partial\gamma} &= -\frac 1 2 \sum_{k=1}^N \left\{ \frac{\partial\lambda_k}{\partial\gamma}\lambda_k^{-1} - \frac{\partial\lambda_k}{\partial\gamma} \lambda_k^{-2}\left(\sum_{i=1}^N u_{k}^i\vx_i\cdot\vsa\right)^2 \right. \\
		&+ \left. 2\lambda_k^{-1} \left(\sum_{i=1}^N u_k^i\vx_i\cdot\vsa\right) \left(\sum_{i=1}^N \frac{\partial u_k^i}{\partial\gamma} \vx_i\cdot\vsa\right) \right\}, 
	\end{split}
	\label{eq:gradlikelihood_3}
\end{equation}
The derivatives of $\vec{u}_k(\rho,\gamma)$ and $\lambda_k(\rho,\gamma)$ with respect to $\rho$ can then be computed using the following equations~\cite{matrix_cook_book}: 
\begin{equation}
	\frac{\partial\lambda_i(\rho,\gamma)}{\partial\rho} = \vec{u}_k(\rho,\gamma)^T\frac{\partial\bm{G}(\rho,\gamma)}{\partial\rho}\vec{u}_k(\rho,\gamma)
	\label{eq:grad_eigval}
\end{equation}
and
\begin{equation}
	\frac{\partial \vec{u}_k(\rho,\gamma)}{\partial\rho} = \sum_{l\neq k} \left( \vec{u}_k(\rho,\gamma)^T\frac{\partial\bm{G}(\rho,\gamma)}{\partial\rho}\vec{u}_l(\rho,\gamma) \right) \left(\lambda_k(\rho,\gamma) - \lambda_l(\rho,\gamma)\right)^{-1} \vec{u}_l(\rho,\gamma).
	\label{eq:grad_eigvec}
\end{equation}
Equivalent equations can be written for their derivatives with respect to $\gamma$.

@@

In order to evaluate the accuracy of our inference procedure, we want to estimate how close the covariance matrix reconstructed by likelihood maximization $\bm{C}_{max}$ is to the real covariance matrix $\bm{C}$. 
To establish  a reference point for inference quality, we have to compare $\bm{C}$ to the empirical covariance matrix $\bm{C}_{emp}$ computed as if observations were independent, \emph{i.e.} $\bm{C}_{emp}(a,b) = N^{-1}\sum_{i=1}^N x_i^a x_i^b$. 
Note that it is also possible to estimate the accuracy of the method by comparing the coupling matrices obtained by inverting each covariance matrix, namely $\bm{J}$, $\bm{J}_{max}$ and $\bm{J}_{emp}$. 
Here, we described results obtained on artificial data.

Data is generated according to the process described in section \ref{sec:statement_of_the_problem}. 
We first build a  balanced binary tree $\mathcal{T}$ with $N=2^9=512$ leaves. 
The length of each branch of $\mathcal{T}$ is chosen from a uniform distribution in the interval $[0,1]$. 
We also sample positive semi-definite coupling matrix $\bm{J}$ of size $L\times L$ with $L=4$ or $L=10$, with entries normally distributed with mean $\mu_J=0.8$ and $\sigma_J=0.2$.

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/PPVs_L4.pdf}
	\caption{Quality of prediction of interactions for  different values of $\gamma$ and system size $L=4$. Interactions are defined as non-zero elements of the coupling matrix. In the $L=4$ case, there are $6$ possible interactions. Predictions are made by taking the largest elements (in absolute terms) of the inferred coupling matrix. The PPV is the fraction of correctly predicted contacts for a given number of predictions. The perfect prediction is to the one that would be made using the real coupling matrix $\bm{J}$.}
	\label{fig:PPV_L4}
\end{figure}

@@

\clearpage
\bibliographystyle{unsrt}
\bibliography{bib_phylo}
